Clinical Phenotype and Survival Analysis in Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies Due to Bi-allelic Loss of Function Variants in \emph{SMPD4}

### 2022-09-09 paper/paper.tex
more edits

\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{caption}
\usepackage{longtable}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Statistical Analysis of Clinical Phenotype and Patient Survival in Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies},
  pdfauthor={Dean Marchiori},
  pdfkeywords={NEDMABA, SMPD4},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{Statistical Analysis of Clinical Phenotype and Patient Survival
in Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and
Structural Brain Anomalies}
\author[]{Dean Marchiori%
\corref{cor1}%
}
 \ead{deanmarchiori@gmail.com}

@@

\begin{abstract}
A recently described, rare genetic condition known as Neurodevelopmental
Disorder with Microcephaly, Arthrogryposis, and Structural Brain
Anomalies (NEDMABA) has been identified in children with bi-allelic
loss-of-function variants in \emph{SMPD4}. The progression of this
condition is not well understood with the limited case reports described
so far exhibiting a severe and clinically diverse phenotype. A gap
exists in the understanding of associations present in the heterogenous
features of the clinical phenotype, and the expected survival
probabilities of affected individuals. This is driven in part to the
paucity of analysis-ready data on reported cases. This research aims to
collate and standardise available case reports, to analyse and identify
meaningful clusters in the clinical phenotype, and to quantify the
survival probability for children with NEDMABA. To overcome the
challenge of multidimensional data on very few subjects, we employ
Multiple Correspondence Analysis (MCA) as a dimension reduction
technique, which is then subject to cluster analysis and interpretation.
To quantify more accurate survival probabilities, Kaplan-Meier
estimation is formulated to account for censoring in the data. The
analysis correctly identified the classic phenotype for this condition,
as well as one other distinct feature cluster relating to findings of
vocal cord paralysis and respiratory failure. Several additional
outlying and isolated clinical features were also identified. The
survival probability for those affected was found to decline sharply in
early infancy with median survival of 150 days, but with some surviving
as long as 12.5 years. This wide range of outcomes is provisionally
associated with variant type however interpretation of these results are
guarded based on very low sample sizes. This analysis represents the
first of its kind to help describe associations and trajectories of
individuals with this condition, despite challenges with sparse and
inconsistent data. This analysis can provide clinicians and genetic
counsellors with better information to aide in decision making and
support for families with this rare condition.
\end{abstract}

\begin{keyword}
    NEDMABA \sep 
    SMPD4
\end{keyword}
\end{frontmatter}\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, breakable, sharp corners, interior hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, frame hidden]}{\end{tcolorbox}}\fi

@@

Further analysis to better describe rare genetic conditions was explored
in \citet{diaz2020phenotype} in the context of large scale
genotype-phenotype analysis. The authors argue patients with rare
disorders (often in small samples) often present with varied symptoms
that do not match exactly with the described phenotype. When considering
this situation of low sample size data with many features, methods in
the field of multivariate statistical analysis are commonly deployed.
Here the aim is to find substructure in the data, or a simple
representation of the multi-dimensional space. Methods, such as Multiple
Correspondence Analysis \citep{le2010multiple} are appropriate for high
dimensional data comprised of binary indicators. This technique is cited
in many studies in the application of dimension reduction and clustering
of comorbidities and phenotypes from disease
\citep{han2018cluster, costa2013use}.

@@

In the case where two variables were formed from features that are
considered to be synonymous, these were merged into one indicator column
to prevent duplication e.g.~\{bilateral\_cleft\_lips,
bilateral\_cleft\_lip, cleft\_lip\_b\_l\} -\textgreater{}
\{bilateral\_cleft\_lip\}.

@@

This format was preserved and other case studies identified in the
literature
\citep{ravenscroft2021neurogenetic, monies2019lessons, bijarnia2022growth, ji2022case}
were manually entered to conform to this template to allow the data to
be combined for further analysis.

@@

The data on subjects introduced above is subset to include only
variables that can be transformed into binary indicator variables that
represent the presence or absence of a given clinical feature. Only
those features that appeared in more than one case were included to
minimise the influence of isolated features in this analysis.

Dimension reduction on this wide data set was performed using Multiple
Correspondence Analysis (MCA) (an analogy to Principle Component
Analysis (PCA) for categorical data) \citep{le2010multiple} in order to
arrive at a set of meaningfully small dimensions that account for most
of the variation in the data.

@@

This computation of the Multiple Correspondence Analysis was computed
using the \emph{FactoMineR} R package \citep{factominer}.

@@

A range of clustering algorithms were compared including AGNES
hierarchical agglomerative \citep{kaufman2009finding}, k-means
\citep{hartigan1979algorithm} and self-organising maps (SOM)
\citep{kohonen2012self}. A sensible range of clusters from 2-5 were
trialed for each clustering algorithm. The results for each method were
evaluated using three measures of stability: Average proportion of
non-overlap (APN), Average Distance (AD) and Average Distance Between
Means (ADM) \citep{datta2003comparisons}. Stability measures compare the
full clustering result with a result based on dropping each of the
columns one at a time. These metrics were selected to represent each
data point (clinical feature) with a number of MCA dimensions. A good
cluster solution is one that is robust and meaningful across all of
these dimensions Figure~\ref{fig-clvalid}. The optimal method and
cluster number was selected based on a trade off between the best
performing method using the evaluation metrics and a sensible
partitioning of the data for this analysis Table~\ref{tbl-eig}.

@@

The initial results of the MCA were inspected with a bi-plot and
identified two subjects (17, 18) from the data that were clear outliers
and sat significantly beyond the 95\% confidence ellipse. These
individuals represented two twins from \citet{magini2019loss} who were
described with a much milder phenotype and were subject to a missense
variation in \emph{SMPD4}. Further analysis of this influence is
conducted below. For now, these subjects were removed to allow for more
meaningful analysis.

@@

The \emph{contribution} of the categories were ranked within each of the
first four MCA dimensions (Figure~\ref{fig-ctr}). The first dimension is
dominated by highly specific facial dysmorphisms such as epicanthal
folds, short philtrum and low anterior hairline.

The second MCA dimension is categorised by feeding and respiratory
dysfunction with vocal cord palsy, tracheostomy, tube feeding and GERD.

Examining the bi-plot of the first two principal axes
Figure~\ref{fig-biplot} illustrates the contrasting sources of variation
between conditions relating to airway and swallowing function and
predominantly facial features and structural brain anomalies such as as
cerebellar hypoplasia, development delay.

The third dimension is characterised by atypical, complex features such
as patent ductus arteriosus, respiratory infections, small for
gestational age and further facial features. These are contrasted by the
fourth principal dimension which is strongly categorised by respiratory
distress including stridor and hypercapnia in the absence of typical
presentations of simplified gyral pattern, intrauterine growth
retardation and microcephaly.

@@

Cluster validation measures resulted in either 2 or 5 clusters as the
optimal choice.

In order to partition the data into meaningful groupings, k-means with 5
clusters was selected. This reflects the optimal choice for both AD and
FOM cluster metrics.

A scatter plot of the cluster results Figure~\ref{fig-clust} projected
on just the first two dimensions highlights a core cluster of classic
features of NEDMABA such as microcephaly, hypomyelination,
arthrogryposis and feet deformity and structural brain anomalies. A
separate cluster is formed of airway and feeding related conditions
involving vocal cord paralysis, tracheostomy, tube feeding and enlarged
ventricles.

Three other smaller satellite clusters exists where highly distinctive
features were associated with a single individual or family. A full
articulation of feature to cluster assignment is given in
Table~\ref{tbl-profile}.

@@

A comparative box plot of survival times by variant type shows
significant outliers for missense+splice and missense variations.
However, even within missense homozygous variants the survival time
varies greatly. Other missense variants, such as heterozygous missense
with gene deletion do not show the same increased survival time as noted
by \citet{magini2019loss}.

@@

A log-rank test was carried out to compare the survival curves of the
various gene variants and showed a significant difference in the
stratified survival curves (\(\chi^2 = 14.8, p = 0.04)\) however due to
the small sample size, several variants had expected counts less than 5
which would create a violation in assumptions for the test.

@@

The use of Multiple Correspondence Analysis confirmed the heterogeneous
variation in the data with the first four principal axes being required
to explain more than 50\% of the variation in the data set. The
examination of key contributing factors, as well as a visual inspection
of the data projected into this lower dimensional space indicate the
primary sources of variation in the clinical phenotype are explained by
a few individuals with very different clinical features. This has
confounded the typical or classical presentation.

The cluster analysis does not aim to group individuals that are alike,
but rather the clinical features themselves and how these contrasting
clusters account for variation in the data. This highlights common
sources of variation in the clinical phenotype of the condition itself,
rather than typical coincident features of any one child.

After excluding two outlying, remarkable cases this analysis clearly
identifies a core cluster of the hallmark features of microcephaly,
hypomyelination, arthrogryposis, simplified gyral pattern and
developmental delay. Of significant note is the presence of an adjacent
cluster of related features that are not as well described in The Human
Phenotype Ontology \citep{kohler2021human} as shown in
Table~\ref{tbl-hpo}. We identify a prolific feature of `respiratory
distress' as an expression more associated with other features such as
vocal cord palsy, tracheostomy, tube feeding and GERD which are present
in the case reported data, but not well described in HPO. This
represents a related but distinct grouping of severe features associated
with this condition.

Other smaller clusters identify outlying features that are expressed in
individuals or specific families. Many of the outlying features are
highly specific facial or anatomical dysmorphisms. While coarse facial
features are a noted and core presentation, those with a milder course
had clinical reports that articulated these dysmorphisms in much more
detail. On a similar note, the reporting of `talipes' as a condition
which seems distinct as a source of variation, appearing as an isolated
feature without the typical microcephaly and hypomyelination. However,
there was significant heterogeneity in how feet contractures were
reported in clinical case reports. In the data pre-processing, a
separate feature was created for reported cases with rocker bottom feet
or congenital vertical talus versus those with clubfeet/talipes. In
practice its not clear how these diagnoses were arrived at, and given
the non-specific nature of these features, care needs to be taken in
distinguishing between similar manifesting features in the literature
where possible. This is particularly clear as rocker bottom feet was
found closely related in our analysis to other core features despite the
prevalence in HPO being quite low (2/21), suggesting possible
under-reporting or confounded reporting of this feature. Another feature
that clusters within the core identified features is bilateral cleft
lip. This condition has a reported frequency of just 2/21
(Table~\ref{tbl-hpo}) and may be an emerging feature for a more typical
presentation given its close relationship with other classic features.
Not all features in the data were included in this analysis, particular
if isolated to just one case. In this circumstance, these terms in
Table~\ref{tbl-hpo} have been labelled as `Isolated'.

While the number of clusters chosen all represent valid partition of the
data, some metrics indicated an optimal selection of 2 clusters. This
type of partition is useful when segregating outliers from the rest of
the data and is valid in some applications. In this study, a larger
number was selected to align the goals of the analysis which was to
uncover distinct groups of features.

In terms of patient survival, the first few months show a sharp decline
in survival probability confirming early demise in the infant period as
a key factor. A small number of cases show much longer survival into
childhood and in some cases into the second decade of life. This work
identifies a one year survival rate of 0.403 (95\% CI: 0.252, 0.644). A
naive estimate, using just the number of deceased patients at this time
as proportion of the study population is 0.57. This highlights the
importance of applying a survival analysis technique that incorporates
censoring so as not to erroneously overestimate the survival
probability. The current literature suggests the type of gene variant
has an impact on survival time, which seems to be supported in this
work. Through a comparison of multiple survival curves, stratified by
variation type, the limited sample size prevented other meaningful
analysis such as log-rank tests or cox proportional hazard modelling
which are likely under-powered when comparing multiple groups.

Despite NEDMABA exhibiting a diverse and heterogeneous clinical
phenotype, our analysis was able to identify meaningful substructures in
this data through the use of dimension reduction techniques (MCA) and
cluster analysis. This has isolated the classic NEDMABA phenotype of
microcephaly, hypomyelination, arthrogryposis, simplified gyral pattern
and developmental delay. Furthermore it has isolated additional distinct
groupings of features that are exhibited in patients such as vocal cord
paralysis with tracheostomy, swallowing dysfunction, tube feeding and
GERD. Furthermore, we were able to quantify the expected baseline
survival probabilities for children with this condition as well as
quantifying the significant variation in these estimates. Future work is
required with the benefit of more cases, to better understand the
differences in survival curves for other factors associated with
genotype.

>>>>

\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{caption}
\usepackage{longtable}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Clinical Phenotype and Survival Analysis in Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies Due to Bi-allelic Loss of Function Variants in SMPD4},
  pdfauthor={Dean Marchiori},
  pdfkeywords={NEDMABA, SMPD4},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{Clinical Phenotype and Survival Analysis in Neurodevelopmental
Disorder with Microcephaly, Arthrogryposis, and Structural Brain
Anomalies Due to Bi-allelic Loss of Function Variants in \emph{SMPD4}}
\author[]{Dean Marchiori%
\corref{cor1}%
}
 \ead{deanmarchiori@gmail.com}

@@

\begin{abstract}
A recently described, rare genetic condition known as Neurodevelopmental
Disorder with Microcephaly, Arthrogryposis, and Structural Brain
Anomalies (NEDMABA) has been identified in children with bi-allelic
loss-of-function variants in \emph{SMPD4}. The progression of this
condition is not well understood with the limited case reports described
so far exhibiting a severe and clinically diverse phenotype. A gap
exists in the understanding of associations present in the heterogenous
features of the clinical phenotype, and the expected survival
probabilities of affected individuals. This is driven in part to the
paucity of analysis-ready data on reported cases. This research aims to
collate and standardise available case reports into a common dataset, to
analyse and identify meaningful clusters in the clinical phenotype, and
to quantify the survival probability for children with NEDMABA. To
overcome the challenge of multidimensional data on very few subjects, we
employ Multiple Correspondence Analysis (MCA) as a dimension reduction
technique, which is then subject to cluster analysis and interpretation.
To account for censoring in the data Kaplan-Meier estimation is
formulated calculate patient survival time. The analysis correctly
detected the classic phenotype for this condition, as well as an
additional distinct feature-cluster relating to findings of vocal cord
paralysis, feeding dysfunction and respiratory failure. Several
additional outlying and isolated clinical features were also identified.
The survival probability for those affected was found to decline sharply
in early infancy with median survival of 150 days, but with some
surviving as long as 12.5 years. This wide range of outcomes is
provisionally associated with different variant types however this
conclusion could not be validated based on very low sample sizes. This
analysis represents the first of its kind to help describe associations
and trajectories of individuals with this newly reported condition,
despite challenges with sparse and inconsistent data. This analysis can
provide clinicians and genetic counsellors with better information to
aide in decision making and support for families with this rare
condition.
\end{abstract}

\begin{keyword}
    NEDMABA \sep 
    SMPD4
\end{keyword}
\end{frontmatter}\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, interior hidden, sharp corners, enhanced, boxrule=0pt, breakable, frame hidden]}{\end{tcolorbox}}\fi

@@

Further analysis to better describe rare genetic conditions was explored
in \citet{diaz2020phenotype} in the context of large scale
genotype-phenotype analysis. The authors argue patients with rare
disorders (often in small samples) frequently present with varied
symptoms that do not match exactly with the described phenotype. When
considering this situation of low sample size data with many features,
methods in the field of multivariate statistical analysis are commonly
deployed. Here the aim is to find substructure in the data, or a simple
representation of the multi-dimensional space. Methods, such as Multiple
Correspondence Analysis \citep{le2010multiple} are appropriate for high
dimensional data comprised of binary indicators. This technique is cited
in many studies in the application of dimension reduction and clustering
of comorbidities and phenotypes from disease
\citep{han2018cluster, costa2013use}.

@@

In the case where synonymous terms are present, these indicator columns
are consolidated and merged into one indicator column to prevent
duplication e.g.~\{bilateral\_cleft\_lips, bilateral\_cleft\_lip,
cleft\_lip\_b\_l\} -\textgreater{} \{bilateral\_cleft\_lip\}.

@@

This format was used as a template to align other written case reports
identified in the literature
\citep{ravenscroft2021neurogenetic, monies2019lessons, bijarnia2022growth, ji2022case}.
These were manually entered to conform to this template to allow the
data to be combined for further analysis.

@@

The data on subjects introduced above is subset to include only
variables that can be transformed into binary indicator variables that
represent the presence or absence of a given clinical feature. Only
those features that appeared in more than one case report were included
to minimise the influence of isolated features in this analysis.

Dimension reduction on this wide data set was performed using Multiple
Correspondence Analysis (MCA) (an analogy to Principle Component
Analysis (PCA) for categorical data) \citep{le2010multiple} in order to
arrive at a set of meaningfully small dimensions that account for most
of the variation in the data. This computation was carried out using the
\emph{FactoMineR} R package \citep{factominer}.

@@

@@

A range of clustering algorithms were compared including AGNES
hierarchical agglomerative \citep{kaufman2009finding}, k-means
\citep{hartigan1979algorithm} and self-organising maps (SOM)
\citep{kohonen2012self}. A sensible range of clusters from 2-5 were
trialed for each clustering algorithm. The results for each method were
evaluated using three measures of stability: Average proportion of
non-overlap (APN), Average Distance (AD) and Average Distance Between
Means (ADM) \citep{datta2003comparisons}. Stability measures compare the
full clustering result with a result based on dropping each of the
columns one at a time. These metrics were selected to represent each
data point (clinical feature) with a number of MCA dimensions. A good
cluster solution is one that is robust and meaningful across all of
these dimensions (Figure~\ref{fig-clvalid}). The optimal method and
cluster number was selected based on a trade off between the best
performing method using the evaluation metrics and a sensible
partitioning of the data for this analysis (Table~\ref{tbl-eig}).

@@

The initial results of the MCA were inspected with a bi-plot and
identified two subjects (17, 18) from the data that were clear outliers
and sat significantly beyond the 95\% confidence ellipse
(Figure~\ref{fig-biplotprelim}). These individuals represented two twins
from \citet{magini2019loss} who were described with a much milder
phenotype and were subject to a missense variation in \emph{SMPD4}.
Further analysis of this influence is conducted below. For now, these
subjects were removed to allow for more meaningful analysis of other
features.

@@

The \emph{contribution} of the categories were ranked within each of the
first four principal dimensions in Figure~\ref{fig-ctr}. The first
dimension is dominated by isolated cases with highly specific facial
dysmorphisms such as epicanthal folds, short philtrum and low anterior
hairline. The second MCA dimension is categorised by feeding and
respiratory dysfunction with vocal cord palsy, tracheostomy, tube
feeding and GERD. Examining the bi-plot of the first two principal axes
(Figure~\ref{fig-biplot}) illustrates the contrasting sources of
variation between conditions relating to airway and swallowing function
and predominantly facial features and structural brain anomalies such as
as cerebellar hypoplasia, development delay. The third dimension is
characterised by atypical, complex features such as patent ductus
arteriosus, respiratory infections, small for gestational age and
further facial features. These are contrasted by the fourth principal
dimension which is strongly categorised by respiratory distress
including stridor and hypercapnia in the absence of typical
presentations of simplified gyral pattern, intrauterine growth
retardation and microcephaly.

@@

Cluster validation measures resulted in either 2 or 5 clusters as the
optimal choice (Table~\ref{tbl-opticlust}). In order to partition the
data into meaningful groupings, k-means with 5 clusters was selected.
This reflects the optimal choice for both AD and FOM cluster metrics. A
scatter plot of the cluster results shown in Figure~\ref{fig-clust} is
projected on the first two principal dimensions highlights a core
cluster of classic features of NEDMABA such as microcephaly,
hypomyelination, arthrogryposis and feet deformity and structural brain
anomalies. A separate and adjacent cluster is formed of airway and
feeding related conditions involving vocal cord paralysis, tracheostomy,
tube feeding and enlarged ventricles. Three other smaller satellite
clusters exists where highly distinctive features were associated with a
single individual or family. A full articulation of feature to cluster
assignment is given in Table~\ref{tbl-profile}.

@@

A comparative box plot of survival times by variant type shows
significant outliers for compound heterozygous \emph{missense + splice}
and homozygous \emph{missense} variants. However, even within homozygous
missense variants the survival time varies greatly. Other missense
variants, such as heterozygous missense with gene deletion do not show
the same increased survival time as noted by \citet{magini2019loss}.

@@

A log-rank test was carried out to compare the survival curves of
patients stratified by the various gene variants and showed a
significant difference in the stratified survival curves
(\(\chi^2 = 14.8, p = 0.04)\) however due to the small sample size and
large number of possible variants, several had expected counts less than
5 which would create a violation in assumptions for the test.

@@

Often, data collected in studies is fit for descriptive purposes but not
for in-depth statistical analysis. This meta-review of current studies
was able to consolidate and make publicly available, data sets that
allow analysis to be conducted on emerging conditions such as NEDMABA.

The use of dimension reduction techniques were shown to be successful in
uncovering relevant and novel substructure in the data. This was
particularly evident with the detection of the accepted typical
phenotype of microcephaly, hypomyelination, arthrogryposis, simplified
gyral pattern and developmental delay.

After excluding two outlying remarkable cases, this analysis clearly
identifies the presence of an adjacent cluster of related features that
are not as well described in The Human Phenotype Ontology
\citep{kohler2021human} as shown in Table~\ref{tbl-hpo}. We identify the
prolific feature of `respiratory distress' as an expression more
commonly associated with \emph{other} features such as vocal cord palsy,
tracheostomy, tube feeding and Gastroesophageal reflux disease (GERD).
These features are present in the case reported data, but not well
described in HPO and help account for the variation seen in the clinical
phenotype in the first few principal dimensions.

Other smaller clusters identify outlying features that are expressed in
individuals or specific families. Many of the outlying features are
highly specific facial or anatomical dysmorphisms. While coarse facial
features are a noted and core presentation, those with a milder course
had clinical reports that articulated these dysmorphisms in much more
detail, which skewed the interpretation.

On a similar note, talipes appears as an isolated feature without the
typical microcephaly and hypomyelination. However, there was significant
heterogeneity in how feet contractures were reported in clinical case
reports. While distinctions were made during data pre-processing between
rocker-bottom feed and talipes, this likely to make intpretation
somewhat difficult. This is particularly clear as rocker-bottom feet was
found closely related in our analysis to other core features despite the
prevalence in HPO being quite low (2/21), suggesting possible
under-reporting or confounded reporting of this feature.

Another feature clusted within core identified features is bilateral
cleft lip. This condition has a reported frequency of just 2/21
(Table~\ref{tbl-hpo}) and may be an emerging feature for a more typical
presentation given its close relationship with other classic features.

The heterogeneous variation in the data was clearly evident however,
with the first four principal axes being required to explain more than
50\% of the variation in the data set. Visual inspection of the data
projected into this lower dimensional space indicate the primary sources
of variation in the clinical phenotype are explained by a few
individuals with very different clinical features. This has confounded
the typical or classical presentation. Not all features in the data were
included in this analysis, particular if isolated to just one case. In
this circumstance, these terms in Table~\ref{tbl-hpo} have been labelled
as `Isolated'.

While the number of clusters chosen all represent valid partition of the
data, some metrics indicated an optimal selection of 2 clusters. This
type of partition is useful when segregating outliers from the rest of
the data and is valid in some applications. In this study, a larger
number (5) was selected to align the goals of the analysis which was to
uncover distinct groups of features.

In terms of patient survival, the first few months show a sharp decline
in survival probability confirming early demise in the infant period as
a key factor. A small number of cases show much longer survival into
childhood and in some cases into the second decade of life. This work
identifies a one year survival rate of 0.403 (95\% CI: 0.252, 0.644). On
the other hand, a naive estimate using just the number of deceased
patients at this time as proportion of the study population is 0.57
(16/28). This highlights the importance of applying a survival analysis
technique that incorporates censoring so as not to erroneously
overestimate the survival probability. The current literature suggests
the type of gene variant has an impact on survival time, which seems to
be supported in this work. Through a comparison of multiple survival
curves, stratified by variation type, the limited sample size prevented
definitive conclusions with many cells having expected counts less than
5.

Despite NEDMABA exhibiting a diverse and heterogeneous clinical
phenotype, our analysis was able to identify meaningful substructures in
this data through the use of dimension reduction techniques (MCA) and
cluster analysis. This has isolated the classic NEDMABA phenotype of
microcephaly, hypomyelination, arthrogryposis, simplified gyral pattern
and developmental delay. Furthermore it has isolated additional distinct
groupings of features that are exhibited in patients such as vocal cord
paralysis, swallowing dysfunction, tube feeding and GERD. Furthermore,
we were able to quantify the expected baseline survival probabilities
for children with this condition as well as quantifying the significant
variation in these estimates. Future work is required with the benefit
of more cases, to better understand the differences in survival curves
for other factors associated with genotype.
